---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RTMBdist

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/janoleko/RTMBdist/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/janoleko/RTMBdist/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The [`RTMB`](https://kaskr.r-universe.dev/RTMB) package enables powerful and flexible statistical modelling with rich random effect structures using automatic differentiation (AD). However, its built-in support for probability distributions is limited to standard cases. `RTMBdist` fills this gap by providing a collection of non-standard, AD-compatible distributions, extending the range of models that can be implemented and estimated with `RTMB`. Most of the distributions implemented in `RTMBdist` allow for automatic simulation and residual calculation by `RTMB`.

The full list of distributions currently available is given in the [List of distributions](https://janoleko.github.io/RTMBdist/articles/distlist.html) vignette.

Feel free to contribute!

## Installation

You can install the development version of `RTMBdist` from [GitHub](https://github.com/) with:

```{r install, eval = FALSE}
remotes::install_github("janoleko/RTMBdist")
```

## Example

```{r library, message=FALSE}
library(RTMBdist)
```

Let's do numerical maximum likelihood estimation with a `gumbel` distribution:

```{r seed, include=FALSE}
set.seed(123)
```

```{r example}
# simulate data
x <- rgumbel(100, location = 5, scale = 2)

# negative log-likelihood function
nll <- function(par) {
  x <- OBS(x) # mark x as the response
  loc <- par[1]; ADREPORT(loc)
  scale <- exp(par[2]); ADREPORT(scale)
  -sum(dgumbel(x, loc, scale, log = TRUE))
}

# RTMB AD object
obj <- MakeADFun(nll, c(5, log(2)), silent = TRUE)

# model fitting using AD gradient
opt <- nlminb(obj$par, obj$fn, obj$gr)

# model summary
summary(sdreport(obj))
```

Through the magic of `RTMB`, we can also immediately simulate new data from the fitted model and calculate residuals:

```{r sim_residuals}
# simulate new data
x_new <- obj$simulate()$x

# calculate residuals
osa <- oneStepPredict(obj, method = "cdf", trace = FALSE)
qqnorm(osa$res); abline(0, 1)
```


